---
title: "taxPPro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{taxPPro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

An R package to propagate annotations for bugphyzz


## Package installation

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install('sdgamboa/taxPPro')
```

## Propagate categorical attribute values

```{r setup, message=FALSE}
library(taxPPro)
library(bugphyzz)
```
### Import data

```{r, message=FALSE}
aer <- physiologies('aerophilicity')[[1]]
table(aer$Evidence)
```

### Propagate upstream with ancestral state reconstruction (ASR)

```{r, message=FALSE, warning=FALSE}
aer_asr <- propagate(aer, max.tax.level = 'genus', direction = 'upstream')
table(aer_asr$Evidence)
```

### Propagate downstream with inheritance

```{r, message=FALSE, warning=FALSE}
aer_inh <- propagate(aer, max.tax.level = 'genus', direction = 'downstream')
table(aer_inh$Evidence)
```

### Propagate with both ASR and inheritance

```{r, message=FALSE, warning=FALSE}
aer_prop <- propagate(aer, max.tax.level = 'genus', direction = 'both')
table(aer_prop$Evidence)
```

## Propagate an attribute with continous values

```{r}
ph <- physiologies('optimal ph')[[1]]
table(ph$Evidence)
```

### upstream

```{r}
ph_upstream <- propagate(ph, 'genus', 'upstream')
table(ph_upstream$Evidence)
```

### downstream

```{r}
ph_downstream <- propagate(ph, 'genus', 'downstream')
table(ph_downstream$Evidence)
```

### ASR + Inheritance

```{r}
ph_prop <- propagate(ph, 'genus', 'both')
table(ph_prop$Evidence)
```

## Session information

```{r}
sessionInfo()
```
