---
title: "taxPPro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{taxPPro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

An R package to propagate annotations for bugphyzz.

# Package installation

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install('sdgamboa/taxPPro')
```

# Steps/outline

This steps are performed for the prepareDataForPropagation function 
(I think I need to change this long names). Changes include formatting,
resolving duplicates, or removing entries that will not be used for propagation.
Also, perform the first of ASR for strains that couln't be mapped to the
data.tree structure.

1. Import bugphyzz data.
2. Remove NAs in the following fields:
  + Parent_NCBI_ID (remove also unknown).
  + NCBI_ID remove also unknown).
  + Evidence (required for ASR/INH).
  + Frequency (required for ASR/INH).
  + Confidence in curation (required for resolving duplicates).
