---
title: "ASR with a taxonomy tree"
subtitle: "All branch lengths are a unit"
output: 
    html_vignette:
        toc: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(taxPPro)
library(bugphyzz)
library(ete3r) # https://github.com/sdgamboa/ete3r
library(phytools)
library(ape)
library(dplyr)
library(tidytree)
library(ggtree)
```

Step 1. Import a dataset with bugphyzz
An attribute with discrete values (aerophilicity):

```{r}
oxygen <- physiologies("aerophilicity")[[1]]
```


```{r}
select_cols <- c(
    "NCBI_ID", "Attribute", "Evidence", "Frequency", "Rank", 
    "Taxon_name"
)
oxygen <- oxygen[!is.na(oxygen$NCBI_ID) & !is.na(oxygen$Rank), select_cols]
glimpse(oxygen)
```

Step 2. Filter by rank
In this case, I’ll use strain

```{r}
oxygen_sp <- oxygen[oxygen$Rank == "strain",]
head(oxygen_sp)
```

Step 3. Get a taxonomy tree
Using the ete3r package: https://github.com/sdgamboa/ete3r


```{r}
my_tree_data <- getTree(unique(oxygen_sp$NCBI_ID))
my_tree <- my_tree_data@phylo
my_tree
```

The metadata of the nodes is also present:

```{r}
my_tree_metadata <- my_tree_data %>% 
    as_tibble()
head(my_tree_metadata)
```

The tree must be rooted and fully solved (dicotomies only):

```{r}
is.rooted(my_tree)

```

```{r}
is.binary(my_tree)
```

# Step 4. Create a 1-0 matrix for each attribute (discrete)

```{r}
attribute_matrix <- attributeMatrix(oxygen_sp, "NCBI_ID", "Attribute")
```

```{r}
head(attribute_matrix)
```

# Step 5. Divide the 1-0 matrix into character vectors (1 vector per attribute)

```{r}
states <- vector("list", ncol(attribute_matrix))
names(states) <- colnames(attribute_matrix)
for (i in seq_along(states)) {
    states[[i]] <- attribute_matrix[,i]
}
states <- states[vapply(states, \(y) sum(y) > 1, logical(1))]
states <- lapply(states, function(x) {
    output <- as.character(x)
    names(output) <- names(x)
    output
    })

statesc <- states[vapply(states, \(y) length(y) > 1 , logical(1))]
lapply(states, head)
```

# Step 6-7. Estimate ASR and convert likelihoods in entries

```{r}
x = createEntriesASR(x = states, phy = my_tree, df = my_tree_metadata)
head(x)
```

# Step 8. Expand dataset and use makeSignatures

Appending entries with appendASR:

```{r}
oxygen$Evidence <- "EXP" # some are empty
new_oxygen <- appendASR(oxygen, x)
```

```{r}
table(oxygen$Rank)
```

```{r}
table(new_oxygen$Rank)
```

```{r}
a = makeSignatures(
    df = oxygen, tax.id.type = "taxname", tax.level = "species",
    Confidence_interval = 0.7, min.size = 10
)
b = makeSignatures(
    df = new_oxygen, tax.id.type = "taxname", tax.level = "species",
    Confidence_interval = 0.7, min.size = 10
)
```

```{r}
lapply(a, length) %>% 
    data.frame()
```

```{r}
lapply(b, length) %>% 
    data.frame()
```

