---
title: "Explore taxid data"
output:
    html_document:
        toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(taxPPro)
library(purrr)
library(dplyr)
library(bugphyzz)
library(tidyr)
# taxizedb::db_download_ncbi(verbose = TRUE, overwrite = TRUE)
```

# NCBI sets

We need a set of taxa in the NCBI to 1) use as a reference to calculate
completeness and 2) to create a tree data structure for propagation.

The sets are based on the options given at the NCBI taxonomy statistics page:
[NCBI Taxonomy Statistics page](https://www.ncbi.nlm.nih.gov/Taxonomy/taxonomyhome.html/index.cgi?chapter=statistics&unclassified=hide&uncultured=hide&period=&from=&to=).

For convenience these sets were downloaded and included in the extdata
directory of the current package (`taxPPro`):

```{r import sets of taxids}
my_ranks <- c('genus', 'species', 'strain')
dir <- system.file('extdata', package = 'taxPPro')
fnames <- list.files(dir, pattern = 'proc.*.txt', full.names = TRUE)
taxids_ls <- vector('list', length(fnames))
for (i in seq_along(taxids_ls)) {
  taxids <- read.table(fnames[i], header = FALSE)[[1]]
  names(taxids_ls)[i] <- sub('^.*/proc_(.*)_[0-9]*\\.txt', '\\1', fnames[i])
  ranks <- taxizedb::taxid2rank(taxids, db = 'ncbi')
  taxids_ls[[i]] <-  data.frame(taxid = taxids, rank = ranks) |> 
    filter(rank %in% my_ranks) 
}
taxids_summary <- taxids_ls |> 
  bind_rows(.id = 'set') |> 
  count(set, rank) |> 
  pivot_wider(
    names_from = rank, values_from = n, values_fill = 0
  ) |> 
  mutate(total = genus + species + strain) |> 
  arrange(total, set)
DT::datatable(
  taxids_summary, 
  caption = 'Table 1. Number of taxa per rank and NCBI set.'
)
```

# Percentage of bugphyzz taxids annotated in the NCBI:

Get a nested list of the sets of NCBI taxids:

```{r}
taxids_ls2 <- taxids_ls |> 
  map(~ split(.x, factor(.x$rank))) |> 
  map_depth(.depth = 2, .f = ~.x$taxid)
str(taxids_ls2, max.level = 2, vec.len = 1)
```

Import bugphyzz physiologies:

```{r}
phys <- physiologies(remove_false = TRUE, full_source = FALSE)
```

Create a list of bugphyzz taxids:

```{r}
bp_ids <- phys |> 
  map(~ pull(filter(.x, NCBI_ID != 'unknown'), NCBI_ID)) |> 
  unlist(use.names = FALSE, recursive = TRUE) |> 
  unique()
bp_ranks <- taxizedb::taxid2rank(bp_ids, db = 'ncbi')
bp_df <- data.frame(taxid = bp_ids, rank = bp_ranks) |> 
  filter(!is.na(bp_ranks))
bp_ids_ls <- split(bp_df, factor(bp_df$rank)) |> 
  {\(y) map(y, ~ .x[[1]])}() |> 
  {\(y) y[my_ranks]}()
str(bp_ids_ls, vec.len = 1)
```

Calculate percentage of bugphyzz taxa in every NCBI set:

```{r}
myFun <- function(taxids_var) {
  names(taxids_var)
  bp_subset <- bp_ids_ls[names(taxids_var)]
  per <- map2_dbl(bp_subset, taxids_var, ~ round(mean(.x %in% .y) * 100))
  data.frame(per) |> 
    tibble::rownames_to_column(var = 'rank')
}
bp_ids_summary <- map(taxids_ls2, myFun) |> 
  bind_rows(.id = 'set') |> 
  pivot_wider(names_from = 'rank', values_from = 'per', values_fill = 0) |> 
  left_join(taxids_summary[,c('set', 'total')], by = 'set') |> 
  arrange(total, set)
DT::datatable(
  bp_ids_summary,
  caption = 'Table 2. Percentage of bugphyzz taxa also annotated in the NCBI sets per rank.'
)
```
# Creating a data.tree stucture based on strains

```{r}
uu_strains <- taxids_ls2$exclude_unclassified_uncultured$strain
uu_strains_class <- taxizedb::classification(uu_strains, db = 'ncbi')
uu_strains_tbl <- map(uu_strains_class, classif2Table) |> 
  bind_rows()
uu_strains_sp <- unique(uu_strains_tbl$species)
map(taxids_ls2, ~ round(mean(uu_strains_sp %in% .x$species) * 100))
```

```{r}
exclude_all_sp <- taxids_ls2$exclude_all$species
sp_children_strain <- taxizedb::children(exclude_all_sp) |> 
  discard(~ !nrow(.x)) |> 
  bind_rows() |> 
  filter(rank == 'strain')
nrow(sp_children_strain)
```

```{r}
mean(bp_ids_ls$strain %in% sp_children_strain$id)
```

```{r}
ust2 <- drop_na(uu_strains_tbl)
```

```{r}
mean(bp_ids_ls$genus %in% ust2$genus)
```


```{r}
mean(bp_ids_ls$species %in% ust2$species)
```

```{r}
mean(bp_ids_ls$strain %in% ust2$strain)
```

Creating a data strucutre based on strains is not a good idea. A good chunk
of genera (from 100 to 63) and species (from 71 to 40) gets dropped.


# Creating a tree based on species



Given the information above, the option of excluding the unclassified and 
uncultured taxa from the NCBI taxonomy gives the most coverage of genera,
species, and strains without adding too many unclassified taxa.

# Session information

```{r}
sessioninfo::session_info()
```

