---
title: "internal_nodes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(phytools)
library(dplyr)
library(tidyr)
library(tibble)
```

## Data

```{r}
data('primate.tree')
data('primate.data')
tre <- primate.tree
dat <- primate.data
states_original <- dat |> 
    rownames_to_column(var = 'tip_label') |> 
    select(tip_label, Activity_pattern) |> 
    mutate(state = 1) |> 
    pivot_wider(
        names_from = 'Activity_pattern', values_from = 'state', values_fill = 0
    ) |> 
    column_to_rownames(var = 'tip_label') |> 
    as.matrix() |> 
    magrittr::set_colnames(LETTERS[1:3])
states_original <- states_original[tre$tip.label,]


macaca <- grep('^Macaca_', rownames(states_original)) # ?
galago <- grep('^Galago_', rownames(states_original)) # B
eulemur <- grep('^Eulemur', rownames(states_original)) # C

states <- states_original
states[] <- 1 / ncol(states)
# m1[macaca, ] <- c(rep(1,3), rep(0, 6))
states[galago, ] <- c(rep(0,4), rep(1, 4), rep(0, 4))
states[eulemur, ] <- c(rep(0,8), rep(1, 4))
myMod <- 'ARD'
myPi <- 'fitzjohn'
```


## Annotate one genus with an additional tip of zero length

This is the recommended way of adding an annotations to an internal node.

```{r}
macaca <- grep('Macaca', rownames(states))
macaca_mrc <- findMRCA(tree = tre, tips = rownames(states)[macaca])
t1 <- bind.tip(tre, tip.label = 'Macaca_gn1', edge.length = 0, where = macaca_mrc)
m1 <- states_original
# m1[] <- 1 / 3
m1 <- rbind(m1, matrix(c(A = 0, B = 0, C = 1), nrow = 1, byrow = TRUE))
rownames(m1)[nrow(m1)] <- 'Macaca_gn1'
m1[macaca,] <- rep(1 / ncol(m1), length(macaca) * 3)

fit1 <- fitMk(
    tree = t1, x = m1, model = myMod, pi = myPi, lik.func = "pruning",
    logscale = TRUE
)
ace1 <- ancr(fit1, tips=TRUE)
plot(ace1, args.plotTree = list(direction = "upwards"))
```

However, in some cases, one might need to add two or more tips carrying
conclicting annotations. If both are of length zero, this throws an error:

```{r}
macaca_mrc2 <- findMRCA(t1, tips = rownames(states)[macaca])
t2 <- bind.tip(t1, tip.label = 'Macaca_gn2', edge.length = 0, where = macaca_mrc2)
m2 <- states
m2 <- rbind(m2, matrix(c(A = 0, B = 0, C = 1), nrow = 1, byrow = TRUE))
m2 <- rbind(m2, matrix(c(A = 0, B = 1, C = 0), nrow = 1, byrow = TRUE))
rownames(m2)[nrow(m2) - 1] <- 'Macaca_gn1'
rownames(m2)[nrow(m2)] <- 'Macaca_gn2'
m2[grep('^Macaca_[^g][^n][^\\d].*', rownames(m2)),] <- rep(1 / ncol(m2), length(macaca) * 3)
fit2 <- tryCatch(
    error = function(e) e, 
    {
        fitMk(
            tree = t2, x = m2, model = myMod, pi = myPi,
            lik.func = "pruning", logscale = TRUE
        )
    }
)
print(fit2$message)
```

## Annotate one genus with an additional tip of length 1e-06

A solultion to the issue described above could be adding small quantity to the
branch length of the additional tips instead of a hard zero, e.g., 1e-06. At
least with the example below, the output seems to be the same (compare with
the first very firt case in this document).

```{r}
macaca <- grep('Macaca', rownames(states))
macaca_mrc <- findMRCA(tree = tre, tips = rownames(states)[macaca])
t1 <- bind.tip(tre, tip.label = 'Macaca_gn1', edge.length = 1e-06, where = macaca_mrc)
m1 <- states_original
# m1[] <- 1 / 3
m1 <- rbind(m1, matrix(c(A = 0, B = 0, C = 1), nrow = 1, byrow = TRUE))
rownames(m1)[nrow(m1)] <- 'Macaca_gn1'
m1[macaca,] <- rep(1 / ncol(m1), length(macaca) * 3)

fit1 <- fitMk(
    tree = t1, x = m1, model = myMod, pi = myPi, lik.func = "pruning",
    logscale = TRUE
)
ace1 <- ancr(fit1, tips=TRUE)
plot(ace1, args.plotTree = list(direction = "upwards"))
```

## Addint two tips with branch length of 1e-06

Now let's see if this has the expected result if I have two annotations.
For this, I'll make all annotations uncertain, except for the tips of the genus 
Eulermur (C/red) and Galago (B/white). I'll add two conflicting annotations
to the genus Macaca using two tips with branch length of 1e-06.
This seems to have given the expected result: half and half.

```{r}
t3 <- bind.tip(tre, tip.label = 'Macaca_gn1', edge.length = 1e-06, where = macaca_mrc)
macaca_mrc3 <- findMRCA(t3, tips = rownames(states)[macaca])
t4 <- bind.tip(t3, tip.label = 'Macaca_gn2', edge.length = 1e-06, where = macaca_mrc3)
m3 <- states
m3 <- rbind(m3, matrix(c(A = 1, B = 0, C = 0), nrow = 1, byrow = TRUE))
m3 <- rbind(m3, matrix(c(A = 0, B = 1, C = 0), nrow = 1, byrow = TRUE))
rownames(m3)[nrow(m3) - 1] <- 'Macaca_gn1'
rownames(m3)[nrow(m3)] <- 'Macaca_gn2'
m3[grep('^Macaca_[^g][^n][^\\d].*', rownames(m3)),] <- rep(1 / ncol(m3), length(macaca) * 3)
fit3 <- fitMk(
    tree = t4, x = m3, model = myMod, pi = myPi, lik.func = "pruning",
    logscale = TRUE
)
ace3 <- ancr(fit3, tips=TRUE)
plot(ace3, args.plotTree = list(direction = "upwards"))
```

## Adding an annotation for one of the descendant species

If I add an annotation for one of the descendat species of the target genus,
the approach that I tried above still works.

```{r}
t3 <- bind.tip(tre, tip.label = 'Macaca_gn1', edge.length = 1e-06, where = macaca_mrc)
macaca_mrc3 <- findMRCA(t3, tips = rownames(states)[macaca])
t4 <- bind.tip(t3, tip.label = 'Macaca_gn2', edge.length = 1e-06, where = macaca_mrc3)
m3 <- states
m3 <- rbind(m3, matrix(c(A = 1, B = 0, C = 0), nrow = 1, byrow = TRUE))
m3 <- rbind(m3, matrix(c(A = 0, B = 1, C = 0), nrow = 1, byrow = TRUE))
rownames(m3)[nrow(m3) - 1] <- 'Macaca_gn1'
rownames(m3)[nrow(m3)] <- 'Macaca_gn2'
m3[grep('^Macaca_[^g][^n][^\\d].*', rownames(m3)),] <- rep(1 / ncol(m3), length(macaca) * 3)
m3['Macaca_nemestrina',] <- c(0, 0, 1)
fit3 <- fitMk(
    tree = t4, x = m3, model = myMod, pi = myPi, lik.func = "pruning",
    logscale = TRUE
)
ace3 <- ancr(fit3, tips=TRUE)
plot(ace3, args.plotTree = list(direction = "upwards"))
```



# Session information

```{r}
sessioninfo::session_info()
```
